The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-01-10 Paper.Rmd
finished (for now) editing the Q1 sections and related materials in supp mat

---
title: "The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W Sheppard, Mark T. Holder, Terrance E. Loecke, P. Chris Reid, Daniel C. Reuman"
date: "December 16, 2018"
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

\begin{figure}
\begin{center}
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption{(A-C) Data generated using the BB1 copula (see section on methods for Q1) 
with three different 
sets of parameters chosen to produce a variety of 
lower-tail (LT) and upper-tail (UT) dependence strengths (methods for Q1), but the same 
Kendall correlation (K) (to two digits) in all cases. (D-F) Similar to A-C, but
with stronger correlation. Asymmetry of tail dependence (LT-UT) is the same 
within each column of panels.\label{fig_cop_pedag}}
\end{center}
\end{figure}
<!--***DAN: Shya, I changed the abbreviation for Kendall Correlation from KC to K in the 
caption, please change in the code that generates the figures, as well
***Shya: Done.-->

@@

<!--bivariate example-->
We present the example of the soil C and N data (see Data), pictured in 
Fig. \ref{fig_biv_multi_raw_cop}A (raw data) and E (copula plot). The variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent
(Fig. \ref{fig_biv_multi_raw_cop}A,E).
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted the 16 copulas
mentioned in the previous section to the 
data pictured in Fig. \ref{fig_biv_multi_raw_cop}E; AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions were computed (Table \ref{SM-tab_soilCNfit}). 
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
normal copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
The model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for model averaging). 
These values are distinct from what a normal copula would give, namely, 0, answering the second part 
of Q1 for these data. The values also differed substantially 
from each other, answering the third part of Q1. The numbers reflect stronger upper- than lower-tail
dependence for the soil C and N data, and this can also be seen visually in the extreme tails 
of the copula plot (Fig. \ref{fig_biv_multi_raw_cop}E).

@@

<!--bivariate summary-->
Table \ref{tab_bivar_summary} summarizes bivariate results for Q1. For all $4$ 
bivariate datasets, the two variables were always significantly related (1$^{st}$ row of 
Table \ref{tab_bivar_summary}). A non-normal copula 
always emerged as the lowest-AIC (highest-AIC-weight) copula 
(2$^{nd}$ row of Table \ref{tab_bivar_summary}). 
The N copula was always a poor fit compared to the highest-AIC-weight copula 
(3$^{rd}$ and 4$^{th}$ rows of Table \ref{tab_bivar_summary}), answering the first part of Q1. 
Upper- and lower-tail statistics (7$^{th}$ and 8$^{th}$ rows of 
Table \ref{tab_bivar_summary}) differed substantially from 0, the tail dependence statistics of 
the N copula, answering the second part of Q1; and also differed substantially from each other
(9$^{th}$ row of Table \ref{tab_bivar_summary}),
answering the third part of Q1. The first three datasets had stronger upper- than lower-tail dependence, and the last dataset had the reverse. 
Although these results convincingly show non-normal copula structure, the 
model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula was a poor fit in some cases 
(5$^{th}$ and 6$^{th}$ rows of Table \ref{tab_bivar_summary}).
But our nonparametric statistics confirmed the same conclusions (10$^{th}$ row of 
Table \ref{tab_bivar_summary}).
Detailed results for each dataset are in Tables \ref{SM-tab_soilCNfit} - \ref{SM-tab_cc_asym_2000} 
and Figs. \ref{SM-fig_soilCN_nonparam} - \ref{SM-fig_cc2000_nonparam}.

We present our analysis of the green spruce aphid abundance data (see Data).
Independence was tested for each pair of the 10 sampling locations, and was 
rejected in every case.
<!--***DAN: Right, Shyamolina? I am assuming that because "Indep" does not 
appear anywhere in Table S11, this is correct, but please confirm or correct.
***Shya: Yes, you are right, all the pairwise sampling locations for aphid data rejected independence test and show copula structures.-->
Model selection among copulas was carried out separately for each pair of locations.
Best-fitting (lowest-AIC) copulas were non-normal
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for the best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substiantially 
larger than those for the N copula (Table \ref{SM-tab_count_AICw_normal}), answering 
the first part of Q1 from the Introduction for these data.
Goodness of fit tests based on a Cramer-von Mises statistic 
(Table \ref{SM-tab_count_p_CvM}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data, and likewise for goodness of fit tests based on a Kolmogorov-Smirnov 
statistic (Table \ref{SM-tab_count_p_KS}), demonstrating that the collection of 
copula families we used was sufficiently broad that the copulas of all pairs of 
locations could be characterized by at least one of them, for these data.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently substantially large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second part of 
Q1 for these data. Apparently these aphid data have greater tail dependence 
(both lower and upper) than would be expected from an N-copula null hypothesis. 
Model-averaged lower-tail dependence statistics minus model-averaged 
upper-tail dependence statistics (Table \ref{SM-tab_count_LTmUT}) were positive 
for all but a few locations pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`), 
indicating that lower-tail dependence across space was much stronger than upper-tail 
dependence. Thus the spatial synchrony of rarity in green spruce aphid is stronger 
than the spatial synchrony of outbreak events. This also answers the third part of Q1.

<!--Non-parametric analysis of tail dependence, green spruce aphid abundance data-->
Nonparametric statistics verified both that tail dependence was distinct from that 
of an N copula for the green spruce aphid abundance data, and that it was asymmetric.
Mean values across all pairs of locations of the statistics 
$\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, $\Dsq_l$, and $\Dsq_u$ were positive and
confidence intervals excluded $0$ (the value for an N copula), confirming tail 
dependence distinct from that of an N copula (Table \ref{SM-tab_count_resamp}). 
Mean values of the statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$
also were always positive and had confidence intervals that excluded $0$ 
(Table \ref{SM-tab_count_resamp}), confirming asymmetry of tail dependence and 
stronger lower- than upper-tail dependence. Each of the statistics 
was plotted against geographic (great-circle) distances between sampling locations, 
across all pairs of sampling locations (Fig. \ref{SM-fig_aphid_count}).

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes some of our results using our
multivariate data. In addition to summaries of the green spruce aphid abundance data results 
elaborated in detail above, the table also summarizes results of analyzing 
leaf-curling plum aphid first flight data, abundance data for the dinoflagellate
\textit{Ceratium furca}, and methan flux data (see Data).

We did model selection and non-parametric analysis with all 20 aphid and 22 plankton species and found our two approaches were consistent to each other. Here, to present our results we chose one species from each category.

We found major occurence of non-independent copulae (row 2, Table \ref{tab_multivar_summary}) out of all pairwise locations (row 1, Table \ref{tab_multivar_summary}) considered. Non-Normal copulae, most of the time, emerged as best-fit based on minimum AIC out of 16 copula family considered (row 3, Table \ref{tab_multivar_summary}) answering the first part of question \hyperref[Intro]{\textbf{(Q1)}}. All four data sets showed Normal copulae were always poor fit in comparison to best-fit non-Normal copulae (compare their average AIC-weights from row 4 and row 5 of Table \ref{tab_multivar_summary}). We considered the "good-fit" location pair based on the p-values (>0.01) obtained from both Cramer-von Mises and Kolmogorov-Smirnov statistics (row 6, Table \ref{tab_multivar_summary}). Some spatial data (green spruce aphid's count, Ceratium furca's count and methane flux) showed stronger lower tail dependence (compare row 7 and row 8 from Table \ref{tab_multivar_summary}, also see row 9) whereas Leaf-plum curling's firstflight data showed stronger upper tail dependence which answered the second and third part of question \hyperref[Intro]{\textbf{(Q1)}}. The spatial synchrony (from model selection) results were also verified by the non-parametric statistic which showed exclusively positive (or negative) confidence interval of the distribution $(\Ps_l - \Ps_u)$ estimated from all location pair for lower (or upper)- tail dependent spatial data (except methane-flux data which had marginal lower quantile). Methane flux data (last column of Table \ref{tab_multivar_summary}) did show that there were non-Normal copula as best fit for most of the location pair but its model-averaged lower tail dependence was not significantly stronger in non-parametric analysis results.

We also carried out the same analyses for abundance and first-flight data for the 18 
aphids species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other 
\textit{Ceratium furca} (results not shown). The conclusions were generally supported that 
non-N copula structure and tail dependence, as well as asymmetry of tail dependence, are 
very common.

@@

knitr::kable(tab_multivar_summary, 
             format="latex", 
             caption.short = "Summary of results for multivariate data sets",
             caption = "Summary of results for multivariate data-sets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only; row 10 was for all location pairs. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green-spruce aphid's count",
                        "Leaf-plum curling aphid's firstflight",
                        "Ceratium furca's count","Methane-flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```

>>>>

---
title: "The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, P. Chris Reid, Daniel C. Reuman"
date: "December 16, 2018"
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

\begin{figure}
\begin{center}
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption{(A-C) Data generated using the BB1 copula (see section on methods for Q1) 
with three different 
sets of parameters chosen to produce a variety of 
lower-tail (LT) and upper-tail (UT) dependence strengths (methods for Q1), but the same 
Kendall correlation (K) (to two digits) in all cases. (D-F) Similar to A-C, but
with stronger correlation. Asymmetry of tail dependence (LT-UT) is the same 
within each column of panels.\label{fig_cop_pedag}}
\end{center}
\end{figure}

@@

<!--bivariate example-->
We present the example of the soil C and N data (see Data), pictured in 
Fig. \ref{fig_biv_multi_raw_cop}A (raw data) and E (copula plot). The variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent
(Fig. \ref{fig_biv_multi_raw_cop}A,E).
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted the 16 copulas
mentioned in the previous section to the 
data pictured in Fig. \ref{fig_biv_multi_raw_cop}E; AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions were computed (Table \ref{SM-tab_soilCNfit}). 
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
normal copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
The model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for model averaging). 
These values are distinct from what a normal copula would give, namely, 0, answering the second part 
of Q1 for these data. The values also differed substantially 
from each other, answering the third part of Q1. The numbers reflect stronger upper- than lower-tail
dependence for the soil C and N data, and this can also be seen visually in extreme upper-right
and lower-left corners of the copula plot (Fig. \ref{fig_biv_multi_raw_cop}E).

@@

<!--bivariate summary-->
Table \ref{tab_bivar_summary} summarizes bivariate results for Q1. For all $4$ 
bivariate datasets, the two variables were always significantly related 
(Table \ref{tab_bivar_summary}, row 1). A non-normal copula 
always emerged as the lowest-AIC (highest-AIC-weight) copula 
(Table \ref{tab_bivar_summary}, row 2). 
The N copula was always a poor fit compared to the highest-AIC-weight copula 
(Table \ref{tab_bivar_summary}, rows 3-4), answering the first part of Q1. 
Upper- and lower-tail statistics (Table \ref{tab_bivar_summary}, rows 7-8) 
differed substantially from 0, the tail dependence statistics of 
the N copula, answering the second part of Q1; and also differed substantially from each other
(Table \ref{tab_bivar_summary}, row 9),
answering the third part of Q1. The first three datasets had stronger upper- 
than lower-tail dependence, and the last dataset had the reverse. 
Although these results convincingly show non-normal copula structure, the 
model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula was a poor fit in some cases 
(Table \ref{tab_bivar_summary}, rows 5-6).
But our nonparametric statistics confirmed the same conclusions 
(Table \ref{tab_bivar_summary}, row 10).
Detailed results for each dataset are in Tables \ref{SM-tab_soilCNfit} - \ref{SM-tab_cc_asym_2000} 
and Figs. \ref{SM-fig_soilCN_nonparam} - \ref{SM-fig_cc2000_nonparam}.

We present our analysis of the green spruce aphid abundance data (see Data)
as an example multivariate analysis.
Independence was tested for each pair of the 10 sampling locations, and was 
rejected in every case.
<!--***DAN: Right, Shyamolina? I am assuming that because "Indep" does not 
appear anywhere in Table S11, this is correct, but please confirm or correct.
***Shya: Yes, you are right, all the pairwise sampling locations for aphid data rejected independence test and show copula structures.
***DAN: Good, thanks, leaving these comments here as a reminder-->
Model selection among copulas was carried out separately for each pair of locations.
Best-fitting (lowest-AIC) copulas were non-normal
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for the best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substiantially 
larger than those for the N copula (Table \ref{SM-tab_count_AICw_normal}), answering 
the first part of Q1 from the Introduction for these data.
Goodness of fit tests based on a Cramer-von Mises statistic 
(Table \ref{SM-tab_count_p_CvM}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data, and likewise for goodness of fit tests based on a Kolmogorov-Smirnov 
statistic (Table \ref{SM-tab_count_p_KS}), demonstrating that the collection of 
copula families we used was sufficiently broad that the copulas of all pairs of 
locations could be characterized by at least one of them, for these data.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently substantially large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second part of 
Q1 for these data. Apparently these aphid data have greater tail dependence 
(both lower and upper) than would be expected from an N-copula null hypothesis. 
Model-averaged lower-tail dependence statistics minus model-averaged 
upper-tail dependence statistics (Table \ref{SM-tab_count_LTmUT}) were positive 
for all but a few locations pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`), 
indicating that lower-tail dependence across space was much stronger than upper-tail 
dependence. Thus the spatial synchrony of rarity in green spruce aphid is stronger 
than the spatial synchrony of outbreak events. This also answers the third part of Q1.

<!--Non-parametric analysis of tail dependence, green spruce aphid abundance data-->
Nonparametric statistics verified that tail dependence was asymmetric for the green 
spruce aphid abundance data.
Mean values across all pairs of locations of the statistics 
$\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, $\Dsq_l$, and $\Dsq_u$ were positive and
confidence intervals excluded $0$ (Table \ref{SM-tab_count_resamp}). 
Mean values of the statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$
also were always positive and had confidence intervals that excluded $0$ 
(Table \ref{SM-tab_count_resamp}), confirming asymmetry of tail dependence and 
stronger lower- than upper-tail dependence. Each of the statistics 
was plotted against geographic (great-circle) distances between sampling locations, 
across all pairs of sampling locations (Fig. \ref{SM-fig_aphid_count}).

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes some of our results using our
multivariate data. In addition to summaries of the green spruce aphid abundance data results 
elaborated in detail above, the table also summarizes results for 
leaf-curling plum aphid first flight data, abundance data for the dinoflagellate
\textit{Ceratium furca}, and methane-flux data (see Data). Results supported the 
conclusions that non-N copula structure, non-N tail dependence, and asymmetric tail 
dependence were common. Most locations pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copula that was not the N copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
substantially better than AIC weights of N copulas (Table \ref{tab_multivar_summary}, rows 4-5).
For these data, best-fitting copulas were nearly always considered an adequate fit, 
according to Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
(e.g., Table \ref{tab_multivar_summary}, row 10). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other 
\textit{Ceratium furca} (results not shown). The conclusions were generally supported that 
non-N copula structure and tail dependence, as well as asymmetry of tail dependence, are 
very common.

@@

knitr::kable(tab_multivar_summary, 
             format="latex", 
             caption.short = "Summary of results for multivariate data sets",
             caption = "Summary of results for multivariate data-sets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only; row 10 was for all location pairs. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green spruce aphid abundance",
                        "Leaf-curling plum aphid first flight",
                        "Ceratium furca abundance","Methane flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```

<!--***DAN: made it to here, 2018 1 10, 1252pm-->
